State of the Practice for Lattice Boltzmann Method Software

### 2021-12-10 StateOfPractice/Papers/LatticeBoltz_SOP-arXiv/LatticeBoltz_SOP.tex
Work in progress on LBM SOP Paper

\documentclass[12pt, notitlepage]{article}
\usepackage{amsmath, mathtools}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{colortbl}
\usepackage{xr}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{xfrac}
\usepackage{tabularx}
\usepackage{float}
\usepackage{siunitx}
\usepackage{booktabs}
\usepackage{caption}
\usepackage{pdflscape}
\usepackage{afterpage}
\usepackage{titling}
\usepackage{blindtext}
\usepackage{multicol}
\usepackage{indentfirst}
\usepackage[toc,page]{appendix}

\usepackage[]{natbib}

%\usepackage{refcheck}

\hypersetup{
    bookmarks=true,         % show bookmarks bar?
      colorlinks=true,       % false: boxed links; true: colored links
    linkcolor=red,          % color of internal links (change box color with linkbordercolor)
    citecolor=green,        % color of links to bibliography
    filecolor=magenta,      % color of file links
    urlcolor=cyan           % color of external links
}

@@

% Used so that cross-references have a meaningful prefix
\newcounter{defnum} %Definition Number
\newcommand{\dthedefnum}{GD\thedefnum}
\newcommand{\dref}[1]{GD\ref{#1}}
\newcounter{datadefnum} %Datadefinition Number
\newcommand{\ddthedatadefnum}{DD\thedatadefnum}
\newcommand{\ddref}[1]{DD\ref{#1}}
\newcounter{theorynum} %Theory Number
\newcommand{\tthetheorynum}{T\thetheorynum}
\newcommand{\tref}[1]{T\ref{#1}}
\newcounter{tablenum} %Table Number
\newcommand{\tbthetablenum}{T\thetablenum}
\newcommand{\tbref}[1]{TB\ref{#1}}
\newcounter{assumpnum} %Assumption Number
\newcommand{\atheassumpnum}{P\theassumpnum}
\newcommand{\aref}[1]{A\ref{#1}}
\newcounter{goalnum} %Goal Number
\newcommand{\gthegoalnum}{P\thegoalnum}
\newcommand{\gsref}[1]{GS\ref{#1}}
\newcounter{instnum} %Instance Number
\newcommand{\itheinstnum}{IM\theinstnum}
\newcommand{\iref}[1]{IM\ref{#1}}
\newcounter{reqnum} %Requirement Number
\newcommand{\rthereqnum}{P\thereqnum}
\newcommand{\rref}[1]{R\ref{#1}}
\newcounter{lcnum} %Likely change number
\newcommand{\lthelcnum}{LC\thelcnum}
\newcommand{\lcref}[1]{LC\ref{#1}}

\usepackage{fullpage}
\usepackage{setspace}
\usepackage{thesis}

\doublespacing

\newcommand{\fulltitle}{State of the Practice for Lattice Boltzmann Method Software}
\newcommand{\authorname}{Peter Michalski}

\begin{document}
\begin{singlespace}

\thesistitle
	{\MakeUppercase{\fulltitle}}
	{\MakeUppercase{\authorname}.}
	{Master of Engineering in Computing and Software}
	{\copyright \ Copyright by \authorname, August 2021}

\newpage
\thispagestyle{empty}

\noindent\textbf{TITLE:} State of The Practice for Lattice Boltzmann Method Software

\noindent\textbf{DEGREE:} Master of Engineering (Computing and Software)

\noindent\textbf{INSTITUTION:} McMaster University

\noindent\textbf{AUTHOR:} Peter Michalski

\noindent\textbf{SUPERVISOR:} Dr. Spencer Smith and Dr. Jacques Carette

\noindent\textbf{NUMBER OF PAGES:} 125

\newpage
\thispagestyle{empty}
\begin{center}
	\textbf{\large Abstract}
\end{center}

We analyze the state of the practice of software development in the Lattice Boltzmann Methods software domain by quantitatively and qualitatively measuring and comparing 23 software packages. We present a general methodology for assessing the state of the practice of software development in scientific computing software domains. We present a domain analysis of the Lattice Boltzmann Methods software family, and identify candidate software packages. We assess these packages to answer software development related research questions to understand how software quality is impacted by software development choices, including principles, processes, and tools. We interview software developers to identify development pain points, and to identify how software quality is ensured. We use quantitative data to rank the software packages using the Analytical Hierarchy Process, and rank Ludwig, ESPResSo, and Palabos as the top three packages. Each of them score high in at least several of the individual qualities that we quantitatively measure. We compare these rankings with rankings from the software development community. We make recommendations, such as providing a detailed user manual and tutorials, explicitly stating the limits of the software, using user-friendly software languages, considering a peer review process, communicating development and contribution information, and using continuous integration and project management tools.\\

\noindent\textbf{Keywords}: Lattice Boltzmann Methods, Scientific Computing, Software Engineering, Software Family, Software Quality

\newpage
\pagenumbering{roman}
\tableofcontents
\addtocontents{toc}{}

~\newpage

\listoffigures

\listoftables

~\newpage

\clearpage
\section*{Reference Material}

\subsection*{Software Engineering Related Definitions and Acronyms}

\noindent\textbf{AHP}: Analytical Hierarchy Process\\

\noindent\textbf{Commonality}: A requirement or goal common to all family members.\\

\noindent\textbf{CPU}: Central Processing Unit\\

\noindent\textbf{Goal}: Goals capture, at different levels of abstraction, the various objectives the system under consideration should achieve \citep{van2001goal}.\\

\noindent\textbf{GPU}: Graphics Processing Unit\\

\noindent\textbf{LOC}: Lines of Code\\

\noindent\textbf{OS}: Operating System\\

\noindent\textbf{OTS}: Off-The-Shelf\\

\noindent\textbf{Requirements}: A software requirement is: \textit{i}) a condition or capability needed by a user to solve a problem or achieve an objective; \textit{ii}) a condition or capability that must be met or possessed by a system or system component to satisfy a contract, standard, specification, or other formally imposed document; or, \textit{iii}) a documented representation of a condition or capability as in the above two definitions \citep{thayer2000ieee}.\\

\noindent\textbf{SCS}: Scientific Computing Software\\

\noindent\textbf{Software Family}: A set of programs with an extensive amount of common properties \cite{parnas1976design}.\\

\noindent\textbf{Variability}: A requirement or goal that varies between family members.

\newpage
\subsection*{Lattice Boltzmann Related Definitions and Acronyms}

\noindent\textbf{1D} 1-Dimensional\\

\noindent\textbf{2D}: 2-Dimensional\\

\noindent\textbf{3D}: 3-Dimensional\\

\noindent\textbf{BGK}: Bhatnagar-Gross-Krook \citep{bhatnagar1954model}\\

\noindent\textbf{CFD}: Computational Fluid Dynamics\\

\noindent\textbf{MRT}: Multi-Relaxation-Time\\

\noindent\textbf{SRT}: Single-Relaxation-Time\\

\noindent\textbf{TRT}: Two-Relaxation-Time\\

\noindent\textbf{LBM}: Lattice Boltzmann Methods\\

\noindent\textbf{LBS}: Lattice Boltzmann Solvers\\

\noindent\textbf{Velocity Directions}: The number of links connecting to each lattice node in the chosen model from neighbouring nodes. All nodes in a chosen lattice model will have the same number of links. A single link will connect between two adjacent nodes.

\end{singlespace}
\newpage
\pagenumbering{arabic}
\section{Introduction}

We analyze the development of Computational Fluid Dynamics (CFD) software packages that use Lattice Boltzmann Methods (LBM). These LBM algorithms consider the behaviours of a collection of particles as a single unit of an intermediate size, and predict their positional probability as they move through a lattice structure. Figure \ref{circularflow} presents the streamlines of converged flow past a stationary circular cylinder with varied Reynolds number, a common task of LBM.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=0.7\textwidth]{circularflow}
		\caption{Streamlines of flow past a stationary circular cylinder at Reynolds number = 10, 20, and 40 \citep{chen2021phase}}
		\label{circularflow}
	\end{center}
\end{figure}

LBM have several advantages over conventional CFD methods, including a simple calculation procedure, improved parallelization, and robust handling of complex geometries \citep{ganji2015application}. LBM are further detailed in Section \ref{domainanalysis}.

Scientific Computing Software (SCS) is defined as the use of computer tools to analyze or simulate mathematical models of real world systems \citep{smith2006systematic}. Given the importance of such software, scientists and engineers desire methods and tools to sustainably develop it with high quality. This has led to the formation of groups like the Software Sustainability Institute (\href{https://www.software.ac.uk/}{SEI}) and Better Scientific Software (\href{https://bssw.io/}{BSS}).

The goal of this report is to analyze the current state of development of LBM SCS to provide insight into its best practices and to offer guidance for future development. We want to understand how software quality in the LBM SCS family is impacted by software development choices, including principles, processes, and tools.

In this report red text denotes internally linked sections of this document. Cyan text denotes an external URL link. Green text denotes a link to the bibliography.

\subsection{Research Questions}\label{purpose}

The LBM software packages are assessed to answer the following research questions:

@@

The packages are assessed for a given set of quality attributes using quantitative and qualitative data. The qualities are defined in Section \ref{softwarequalities}. Quantitative data is measured using quality metrics. The findings are presented in Section \ref{AHPresults}. Qualitative data is gathered from interviews with software package developers. The interview findings are presented in Section \ref{interviewresults}. In this context the purposes of the report are as follows:

\begin{itemize}
	\item Develop and test an updated methodology for assessing the state of the practice of SCS projects
	\item Describe the specific example of LBM software structure and goals
	\item Report on the measure of a subset of LBM software packages along quality metrics
	\item Evaluate the state of the practice of LBM software development along quality attributes
	\item Make suggestions for improving LBM software along quality attributes
	\item Make suggestions for improving state of the practice of SCS assessments
\end{itemize}

\subsection{Motivation}

The purpose of state of the practice assessments is to understand how software quality is impacted by software development choices, including principles, processes, and tools, within SCS communities. This knowledge can be used to guide future development of SCS, specifically along quality attributes, and to reduce software quality failures. This work reports on the state of the practice of LBM software development and makes suggestions on improving the quality of software in this domain.

This assessment of the state of the practice of LBM software development builds off of prior work on assessing the state of research software development. Updates to the methodology used in prior assessments are detailed below. Those assessments include domains such as Geographic Information Systems \citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016}, Oceanographic Software \citep{smith2015state}, Seismology software \citep{SmithEtAl2018}, and statistical software for psychology \citep{SmithEtAl2018_StatSoft}.

In the course of this assessment we updated the methodology that was used during previous assessments. Details of the new methodology are in \citep{methodology2020}. The previous set of research questions were critically assessed and modified. The updated questions are listed in Section \ref{purpose}. In this re-boot we collected more quantitative and qualitative data, focused on software quality attributes, and added more measures, including collecting empirical data and interviewing developers. We have also added a domain analysis to better characterize the functionality provided by the software, and have leveraged the expertise of a domain expert. As in past assessments, the collected data was combined to rank the software using the Analytical Hierarchy Process (AHP). The domain expert was consulted to verify the ordering.

\subsection{Scope}

We analyze a filtered set of LBM software packages along quantitative and qualitative measures. Many of these measures are captured using surface measurements, which can be categorized as initial and easy to capture measurements, of the underlying quality; they may not represent the true quality of the software. Surface measurements are taken as they allow us to apply the same measurement, with reasonable effort, along all software packages despite technical and functional variabilities among the set of software packages. Not all qualities are quantitatively measured. For example, performance is not measured by running the code. A surface investigation of the documentation of each software package for performance information is conducted instead. The measured software packages are open source. Some recommendations may not apply to closed-source software. Practices surrounding close source software development may differ.

The report addresses what was done during the development of the software packages to address the quality attributes listed in Section \ref{softwarequalities}. It then provides general guidance on how to improve these qualities when developing LBM software. It does not make suggestions on what should have been done or should be done for any one specific package. Best practices may differ among software packages due to their inherent organizational and technical differences.

\subsection{Organization}

The report is organized as follows:

\begin{itemize}
	\item \textbf{Introduction} to the report, including its purpose, motivation, scope, and organization.
	\item LBM software \textbf{Domain Analysis}. A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and terminology of a software family \citep{Weiss1997}.
	The purpose of this analysis in this state of the practice assessment is to better classify the different software products based on their functionality.
	\item \textbf{Methodology} of this state of the practice assessment, including the steps of the methodology, software quality definitions and how these qualities are assessed in this report, an overview of candidate software selection, empirical measures gathered, and the AHP. This follows a general methodology for assessing the state of the practice for SCS domains.
	\item \textbf{Quantitative Results}. 
	\item \textbf{Qualitative Results}. 
	\item An analysis of the results and \textbf{Answers To Research Questions}.
	\item \textbf{Conclusion} to the report including comments on suggestions for future state of the practice assessments.
	\item \textbf{Appendix}: This section includes our Research Questions, the Measurement Template, Grading Template, Ethics Approval, and Developer Interview Questions.
\end{itemize}

\newpage

\section{Domain Analysis}\label{domainanalysis}

A domain analysis consists of systematically identifying and documenting the commonalities, variabilities, and parameters of variation of a software family \citep{Weiss1997}. A software family is defined by \cite{parnas1976design} as ``a set of programs whose common properties are so extensive that it is advantageous to study the common properties of the programs before analyzing individual members''.

We added this domain commonality and variability analysis to our original methodology because the first time we did the State of the Practice exercise we found we weren't comparing ``apples to apples". Software packages can have considerable variation, even when the packages appear to be in the same software family. Viewing our final ranking of the software packages as a means for selecting the right tool for a job, functionality is important. A 2D Lattice Boltzmann solver is not a substitute for a 3D Lattice Boltzmann solver when the latter is required. For example, an analysis of a vascular system may require a 3D model to convey all required information to a researcher. This domain analysis helps ensure that the packages that we compare are appropriately similar, and highlights potential differences. This information further helps when analyzing the appropriateness of the final rank of the packages.  This exercise seeks to classify the different software products based on their functionality. Nonfunctional qualities are assessed through other means in our process. For instance, through usability experiments as noted in the \href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/Methodology/Methodology.pdf}{Methodology for Assessing the State of the Practice for Domain X} document. This exercise was not part of the methodology of the previous State of the Practice exercise.

This domain analysis of the LBM SCS family is organized into the four subsection located below. The first subsection reviews the basics of Lattice Boltzmann systems. The next three subsections consist of lists of commonalities, variabilities, and parameters of variation, respectively. These three sections form the heart of the domain analysis and include an extensive set of cross-references to demonstrate the relationships between the different items.

\subsection{Lattice Boltzmann Systems}

LBM are a family of fluid dynamics algorithms for simulating single-phase and multiphase fluid flows, often incorporating additional physical complexities \citep{chen1998lattice} such as reflective and non-reflective boundaries. They consider the behaviours of a collection of particles as a single unit at the mesoscopic scale, between the nanoscopic and microscopic scales. These methods predict the positional probability of a collection of particles moving through a lattice structure. Off-the-shelf (OTS) Lattice Boltzmann Solvers (LBS) allow for a range of fluid and physical model input parameters, computational parameters, and output parameters.

LBS model fluid dynamics within a boundary using a predefined lattice structure and a two step calculation process. The first process is streaming, where the particles move along the lattice via links. The second process is collision, where energy and momentum is transferred among particles that collide \citep{bao2011lattice}.
There are many standardized lattice models - individual solvers within the family might only use a subset of them.
LBM use the initial parameters of the fluid to find the probability of where along the lattice linkages a group of particles are most likely to travel. It then moves the particles into the next node, and transfers the energy and momentum if a collision occurs. Then the process repeats for the duration of the modeling instance.

\subsection{Commonalities}\label{comm_sec}

This section lists common features among potential family members. These features were chosen after a review of LBM literature and software package specifications. The commonalities are organized using the following abstraction of the system, which can be used to describe all Lattice Boltzmann systems: input information, generate the simulation, output the results. Section \ref{comm_lbm} describes the commonalities for the simulation step. Section \ref{comm_in} highlights the input information that is required for Lattice Boltzmann systems. The next section, Section \ref{comm_out}, shows the common features for the output of Lattice Boltzmann systems, such as the requirement that mesh information be written to files. Although the output information could simply be written to the memory, in all practical applications it is desirable to have a persistent record of the output that was created.

Each commonality below uses the same structure. All of the commonalities are assigned a unique item number, which takes the form of a natural number with the prefix ``C''. Following this, a description of the commonality is provided along with a list of related variabilities, which are given as hyperlinks that allow navigation of the document to the text describing the variability.

\subsubsection{Lattice Boltzmann Method Solvers}\label{comm_lbm}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum{}\label{lattice} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & A lattice discretizes a computational domain into a finite number of points. All LBS discretize the computational domain using a regular, evenly spaced grid within a boundary.\\
		\hline
		\bf Related Variability & \vref{numveldir} \vref{varcolop} \vref{varboundary}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{collisionoperator} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & All Lattice Boltzmann versions use a collision operator which concerns collisions between particles. Collision operators map collisions of particles within the lattice space. The Bhatnagar-Gross-Krook Collision Operator is a common LBM collision operator that preserves continuity for a discretized model, for each velocity direction $i$. Its equation is $\mathrm{\Omega_i} = \frac{1}{\tau}(f_{i}^{eq}-f_{i})$, where $\tau$ is the relaxation rate towards equilibrium, $f^{eq}$ is the equilibrium particle probability distribution function, and $f$ is the particle probability distribution function.\\
		\hline
		\bf Related Variability & \vref{dimensions} \vref{numveldir}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{probabilitydensityfunc} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & All Lattice Boltzmann versions use a probability density function to give the probability that fluid has moved into a specific domain.\\
		\hline
		\bf Related Variability & \vref{dimensions}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{equildistfunc} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & Every Lattice Boltzmann version uses an equilibrium distribution function to capture the probability distribution of particles.\\
		\hline
		\bf Related Variability & \vref{varedf} \vref{edfstorage} \vref{edfcoeff}\\
		\hline
	\end{tabular}
\end{minipage}
~\newline
~\newline
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{Boltzmanntransequ} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & Every Lattice Boltzmann version uses a Boltzmann transport equation to describe the statistical behaviour of a system that does not have collisions. The equation is $(\frac{d}{dt} + \mathrm{e} \cdot \nabla_\mathrm{x} + \frac{F}{m} \cdot \nabla_\mathrm{e})f = \Omega(t)$, where $\mathrm{e}$ is the velocity, $\nabla_\mathrm{x}$ is the position vector gradient, $F$ is the force of the fluid, $m$ is the mass, $\nabla_\mathrm{e}$ is the velocity gradient, $f$ indicates the probability density function, and $\Omega(t)$ is the collision operator as a function of time.\\
		\hline
		\bf Related Variability & \vref{vartraneq}\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Input}\label{comm_in}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{fluidandboundaryinfo} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & The LBS require fluid, model, and boundary information for the problem. This includes, but is not limited to, fluid acceleration rate, velocity, and viscosity, as well as the number of dimensions in the lattice model, and the number of velocity directions in the lattice.\\
		\hline
		\bf Related Variability & \vref{interface}\\
		\hline
	\end{tabular}
\end{minipage}\\
\subsubsection{Output}\label{comm_out}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}
		\hline
		\bf Item Number& C\refstepcounter{comnum}\thecomnum\label{predictionstomemory} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colAwidth} | p{\colBwidth}|}		
		\hline
		\bf Description & LBS write fluid predictions, such as the output of the transport equation, to memory.\\
		\hline
		\bf Related Variability & \vref{varvisual} \vref{formatoutput} \\
		\hline
	\end{tabular}
\end{minipage}\\

\subsection{Variabilities}\label{var_sec}
This section provides a list of characteristics that may vary among family members. These features were chosen after a review of LBM literature and software package specifications. As in Section \ref{comm_sec}, the first three subsections on variabilities are organized into the following sublists: Simulation Models, Input and Output. The final subsection lists variabilities that can be characterized as system constraints.

As for the commonalities, each variability is labelled with a unique item number. In this case the numbers are prepended with the letter ``V''. The other three headings provided for each variability are: Description, Related Commonality, and Related Parameter. The related commonalities and parameters are given as a set of identifiers that respectively refer back to the previous section on commonalities or refer forward to the next section on parameters of variation.

\subsubsection{Lattice Boltzmann Method Solvers}
\noindent
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{parallel} \\
		\hline
	\end{tabular}\\

	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}		
		\hline
		\bf Description & LBS may use a framework for parallel processing of the model.\\
		\hline
		\bf Related Commonality & None\\
		\hline
		\bf Related Parameter & \pref{parparallel}\\
		\hline
	\end{tabular}
\end{minipage}\\
~\newline\\
\begin{minipage}{\textwidth}
	\renewcommand*{\arraystretch}{1.5}
	\begin{tabular}{| p{\colCwidth} | p{\colDwidth}|}
		\hline
		\bf Item Number& V\refstepcounter{varnum}\thevarnum\label{varedf} \\
		\hline
	\end{tabular}\\

>>>>

%% 
%% Copyright 2007-2020 Elsevier Ltd
%% 
%% This file is part of the 'Elsarticle Bundle'.
%% ---------------------------------------------
%% 
%% It may be distributed under the conditions of the LaTeX Project Public
%% License, either version 1.2 of this license or (at your option) any
%% later version.  The latest version of this license is in
%%    http://www.latex-project.org/lppl.txt
%% and version 1.2 or later is part of all distributions of LaTeX
%% version 1999/12/01 or later.
%% 
%% The list of all files belonging to the 'Elsarticle Bundle' is
%% given in the file `manifest.txt'.
%% 
%% Template article for Elsevier's document class `elsarticle'
%% with harvard style bibliographic references

%\documentclass[preprint,12pt,authoryear]{elsarticle}
%\documentclass[3p, 12pt,authoryear]{elsarticle}
\documentclass[final, 3p, times, authoryear]{elsarticle}

%% Use the option review to obtain double line spacing
%% \documentclass[authoryear,preprint,review,12pt]{elsarticle}

%% Use the options 1p,twocolumn; 3p; 3p,twocolumn; 5p; or 5p,twocolumn
%% for a journal layout:
%% \documentclass[final,1p,times,authoryear]{elsarticle}
%% \documentclass[final,1p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,3p,times,authoryear]{elsarticle}
%% \documentclass[final,3p,times,twocolumn,authoryear]{elsarticle}
%% \documentclass[final,5p,times,authoryear]{elsarticle}
%% \documentclass[final,5p,times,twocolumn,authoryear]{elsarticle}

%% For including figures, graphicx.sty has been loaded in
%% elsarticle.cls. If you prefer to use the old commands
%% please give \usepackage{epsfig}

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsmath,amsthm}

%% The lineno packages adds line numbers. Start line numbering with
%% \begin{linenumbers}, end it with \end{linenumbers}. Or switch it on
%% for the whole article with \linenumbers.
%% \usepackage{lineno}

\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pgfplots}
\usepackage{arydshln}
\usepackage{paralist}
\usepackage{multicol}

%% Comments
\newif\ifcomments\commentstrue

\ifcomments
\newcommand{\authornote}[3]{\textcolor{#1}{[#3 ---#2]}}
\newcommand{\todo}[1]{\textcolor{red}{[TODO: #1]}}
\else
\newcommand{\authornote}[3]{}
\newcommand{\todo}[1]{}
\fi

\newcommand{\wss}[1]{\authornote{blue}{SS}{#1}} %Spencer Smith
\newcommand{\jc}[1]{\authornote{red}{JC}{#1}} %Jacques Carette
\newcommand{\zm}[1]{\authornote{magenta}{MN}{#1}} %Zahra Motamed
\newcommand{\pmm}[1]{\authornote{cyan}{AD}{#1}} %Peter Michalski

\newcommand{\notdone}[1]{\textcolor{red}{#1}}
\newcommand{\done}[1]{\textcolor{black}{#1}}

@@

\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
+}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}

\journal{???}

\begin{document}

\begin{frontmatter}

%% Title, authors and addresses

%% use the tnoteref command within \title for footnotes;
%% use the tnotetext command for theassociated footnote;
%% use the fnref command within \author or \affiliation for footnotes;
%% use the fntext command for theassociated footnote;
%% use the corref command within \author for corresponding author footnotes;
%% use the cortext command for theassociated footnote;
%% use the ead command for the email address,
%% and the form \ead[url] for the home page:
%% \title{Title\tnoteref{label1}}
%% \tnotetext[label1]{}
%% \author{Name\corref{cor1}\fnref{label2}}
%% \ead{email address}
%% \ead[url]{home page}
%% \fntext[label2]{}
%% \cortext[cor1]{}
%% \affiliation{organization={},
%%            addressline={}, 
%%            city={},
%%            postcode={}, 
%%            state={},
%%            country={}}
%% \fntext[label3]{}

\title{State of the Practice for Lattice Boltzmann Method Software}

%% use optional labels to link authors explicitly to addresses:
%% \author[label1,label2]{}
%% \affiliation[label1]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}
%%
%% \affiliation[label2]{organization={},
%%             addressline={},
%%             city={},
%%             postcode={},
%%             state={},
%%             country={}}

\author[CAS]{Spencer Smith}
\author[CAS]{Peter Michalski}
\author[CAS]{Jacques Carette}
\author[ME]{Zahra Motamed}

\affiliation[CAS]{organization={McMaster University, Computing and Software
Department},%Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

\affiliation[ME]{organization={McMaster University, Mechanical Engineering},%Department and Organization
            addressline={1280 Main Street West}, 
            city={Hamilton},
            postcode={L8S 4K1}, 
            state={Ontario},
            country={Canada}}

\begin{abstract}

    We analyze the state of software development practice for Lattice Boltzmann
    solvers by quantitatively and qualitatively measuring and comparing 24
    software packages for 10 software qualities (installability, correctness/
    verifiability, reliability, robustness, usability, maintainability,
    reusability, understandability, visibility/transparency and
    reproducibility). Our reproducible analysis method employs a measurement
    template (containing 103 measures that are manually and automatically
    extracted from the project repository) and developer interviews (consisting
    of a set of 20 questions). From the measurement results, we ranked the
    software using the Analytic Hierarchy Process (AHP), a multi-criteria
    decision making method appropriate for cases with a mix of qualitative and
    quantitative factors. Our ranking was roughly consistent with ranking the
    repositories by the number of GitHub stars, although the number one project
    by star count (Sailfish) is ranked 9th by our methodology. Our top three
    packages were Ludwig, ESPResSo, and Palabos (all in the top 10 by star
    count). We interviewed 4 developers to gain insight into their current pain
    points. Identified challenges include lack of development time, lack of
    funding, and difficulty with ensuring correctness.  Based on our assessment,
    we make the following recommendations for current and future LBM projects:
    provide a detailed user manual and tutorials, explicitly state the limits of
    the software, use user-friendly programming languages, consider a peer
    review process, communicate development and contribution information, and
    use continuous integration and project management tools.

\end{abstract}

%%Graphical abstract
%\begin{graphicalabstract}
%\includegraphics{grabs}
%\end{graphicalabstract}

%%Research highlights
%\begin{highlights}
%\item Research highlight 1
%\item Research highlight 2
%\end{highlights}

\begin{keyword}
	Lattice Boltzmann Method (LBM), scientific computing, software engineering,
	software quality, Analytic Hierarchy Process
\end{keyword}

\end{frontmatter}

%% \linenumbers

\section{Introduction}

We analyze the development of Computational Fluid Dynamics (CFD) software
packages that use the Lattice Boltzmann Method (LBM). LBM packages form a family
of algorithms for simulating single-phase and multiphase fluid flows, often
incorporating additional physical complexities \citep{chen1998lattice} such as
reflective and non-reflective boundaries. They consider the behaviour of a
collection of particles as a single unit at the mesoscopic scale, between the
nanoscopic and microscopic scales. LBM solvers predict the positional
probability of a collection of particles moving through a lattice structure
following a two step process: i) streaming, where the particles
move along the lattice via links; and, ii) colliding, where energy and
momentum is transferred among particles that collide \citep{bao2011lattice}. As
an example of the output of LBM, Figure~\ref{circularflow} presents the streamlines of
converged flow past a stationary circular cylinder with varying Reynolds numbers.
\wss{Zahra might have a better example?  A journal publication isn't likely
going to accept a figure produced for another paper.}  LBM has several
advantages over conventional CFD methods, including a simple calculation
procedure, improved parallelization, and robust handling of complex geometries
\citep{ganji2015application}.

\begin{figure}[h!]
	\begin{center}
		\includegraphics[width=0.7\textwidth]{./figures/circularflow}
		\caption{Streamlines of flow past a stationary circular cylinder at
		Reynolds number = 10, 20, and 40 \citep{chen2021phase}}
		\label{circularflow}
	\end{center}
\end{figure}

A small sample of important applications of LBM include the following: designing
fuel cells \citep{ZhangEtAl2018}, modelling groundwater flow
\citep{AnwarAndSukop2009}, and analyzing the flow of blood in the cardiovascular
system \citep{SadeghiEtAl2020}.  As these examples illustrate, LBM can be used
for tackling problems that impact such areas as environmental policy,
manufacturing, and health and safety.  Given the important applications of LBM,
users of the libraries will be concerned with software qualities like
reliability, robustness, reproducibility, performance, correctness and
verifiability.  Given the value of their time, developers that create or modify
LBM libraries will have additional quality concerns, including maintainability,
reusability, and understandability.  With the quality considerations in mind,
and the potentially overwhelming number of choices for LBM libraries, the time
is right to assess the state of the practice.  The goal of this report is to
analyze the current state of LBM software development to provide insight into
its best practices and to offer guidance for future development.

To focus our efforts, we devised five research questions, which can be applied
to LBM, as well as to other domains, of research software \citep{SmithEtAl2021}.
The first question below refers to artifacts, which is the term we use to
represent the documents, scripts and code that constitutes a software
development project. Example artifacts include requirements, specifications,
user manuals, unit tests, system tests, usability tests, build scripts, API
(Application Programming Interface) documentation, READMEs, license documents,
process documents, and code.

@@

We investigated the research questions by applying the general methodology
summarized in \citet{SmithEtAl2021}.  The specific application of the
methodology to LBM is reviewed in Section~\ref{methodology}, with the full
details in \citep{Michalski2021}.  The current methodology updates the approach
used in prior assessments of domains like Geographic Information Systems
\citep{SmithEtAl2018_arXivGIS}, Mesh Generators \citep{SmithEtAl2016},
Oceanographic Software \citep{smith2015state}, Seismology software
\citep{SmithEtAl2018}, and statistical software for psychology
\citep{SmithEtAl2018_StatSoft}.  \wss{add citation to medical imaging software,
if it is an option}

With our methodology, we first choose a software domain (in the current case
LBM) and identify a list of 24 software packages.  We approximately measure the
qualities of each package by filling in a grading template. Compared with our
previous methodology, the new methodology also includes repository based
metrics, such as the number of files, number of lines of code, percentage of
issues that are closed, etc.  With the quantitative data in the grading
template, we rank the software with the Analytic Hierarchy Process (AHP). After
this, as another addition to our previous methodology, we interview some of the
development teams to further understand the status of their development process.
Finally, we summarize the measurements (Section~\ref{AHPresults}), summarize the
interview answers (Section~\ref{interviewresults}), answer the research
questions and propose recommendations for LBM software development
(Section~\ref{answersquestions}).

\section{Methodology} \label{methodology}

We developed a methodology for evaluating the state of the practice of
scientific software \citep{SmithEtAl2021}.  This methodology can be instantiated
for a specific domain of scientific software, which in the current case is LBM
software.  The first section below describes the steps of the process we used to
select, measure and compare LBM software. This is followed by quality
definitions and how these qualities were assessed in our project. The rest of
the section provides the details of the methodology.
% an overview of the specifics of candidate software selection, the quantitative
% measurements, developer interviews, the ranking process via AHP and
% interaction with the Domain Expert.
